Tidy data structure to support exploration and modeling of temporal-context data

### 2018-08-16 main.Rmd
done with sec 1.1

## Representing time series and longitudinal data

@@

On the other hand, those who are engaged in longitudinal design, are primarily interested in explaining variations among individuals in change process and making inference about the underlying population. Longitudinal data or panel data typically refers to fewer measurements (small $T$) over a large number of individuals (large $N$). It often occurs to longitudinal design that some individuals are not observed at certain points in time, resulting in an unbalanced panel. This data issue leads representing longitudinal data much more commonly in **long format** where each row holds an observational unit per time and each column contains measurements across the individuals for all points in time. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. less restrictions to be made. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm** [@R-plm].

## Hierarchical and grouped time series

@@

## Tidy data

>>>>

## Time series and longitudinal data

@@

On the other hand, those who are engaged in longitudinal design, are primarily interested in explaining variations among individuals in change process and making inference about the underlying population. Longitudinal data or panel data typically refers to fewer measurements (small $T$) over a large number of individuals (large $N$). It often occurs to longitudinal design that some individuals are not observed at certain points in time, resulting in an unbalanced panel. This data issue leads representing longitudinal data much more commonly in **long format** where each row corresponds to an observational unit per time and each column contains measurements across the individuals for all points in time. Evidently, this data organisation saves storage space for many sparse cells, compared to structuring it in that wide format. Another appealing feature for a long format is data structured in a concise and semantic manner with reference to observations and variables. This subsequently supports perceiving time index as an explicit variable instead of an implicit attribute. In general, fewer restrictions are made if representing temporal-context data in long format.

Longitudinal data representation has been implemented in Stata's time series module [@Stata] and R package **plm** [@plm2008]. The underpinning data structure is a two-dimensional column-homogeneous array. Specifying a longitudinal data set from a generic array needs explicitly declaring time and individuals (or panel variable in Stata's `tsset` command). Therefore, each row of the data can be identified by a specific individual and time point. Individuals, however, can be only declared through a single variable, not multiple.

## Time series of nested and crossed factors

@@

## Tidy data and the grammar of data manipulation
